---
title: "Clean CPSC course data"
author: "Stephan Koenig"
date: 2024-02-05
date-format: medium
embed-resources: true
format:
  html:
    code-tools:
      source: https://github.com/stephan-koenig/clean-cpsc-course-data/blob/main/index.qmd
---

```{r}
#| label: setup
#| message: false

library(conflicted)
library(dplyr)
conflicts_prefer(dplyr::filter)
library(here)
library(janitor)
library(purrr)
library(readxl)
library(stringr)
library(tidyr)
library(writexl)

process_instructor_names <- function(
    multiple_instructors,
    last_and_first_name_combined,
    data) {
  if (multiple_instructors) {
    data <- data |> 
      separate_longer_delim(c(last_name, first_name, rank), delim = regex(";\\s*|/"))
  }
  
  if (last_and_first_name_combined) {
    data <- data |> 
      select(!first_name) |> 
      separate_wider_delim(
        last_name,
        names = c("last_name", "first_name"),
        delim = regex(",\\s*")
      )
  }
  
  data
}

course_counts_file    <- here("data", "raw", "course_counts_raw.xlsx")
undergraduate_courses <- read_xlsx(course_counts_file, sheet = "undergrad")
graduate_courses      <- read_xlsx(course_counts_file, sheet = "grad")
```

## Overview

Using the data in `course_counts_raw.xlsx`:

1)  For courses with multiple instructors (separated by "; ?"), separate into one row for each instructor and instance of a course.

2)  Standardize the names for all persons by

    a)  Fixing the instructor names using consistent capitalization, middle names and spelling.

    b)  Separating first and last name if necessary (separated by "<last>, ?<first>" in the `undergrad` sheet, "<first> <last>" on `grad`).

3)  Calculate the percentage of the course an instructor was responsible for (e.g., for two instructors, it is `0.5`; for three, `0.3333`, etc.).

## Data

### Provenance

`course_counts_raw.xlsx` provided by Alla Sheffer by email on 2024-02-01:

-   SHA256: `{r} cli::hash_file_sha256(course_counts_file)`
-   Origin unknown

### Cleaning

```{r}
#| label: clean-names

undergraduate_courses_clean <- undergraduate_courses |> 
  clean_names() |> 
  rename(
    last_name  = instructor_last_name,
    first_name = instructor_first_name
  ) |> 
  mutate(
    term = as.double(term),
    multiple_instructors = str_detect(last_name, ";"),
    multiple_instructors = replace_na(multiple_instructors, FALSE),
    last_and_first_name_combined = str_detect(last_name, ","),
    last_and_first_name_combined = replace_na(last_and_first_name_combined, FALSE),
    # Add missing rank information
    rank = if_else(
      last_name == "Srinivasan;Wolfman;Feeley",
      "Sessional/Instructor/Instructor",
      rank
    ),
    program = "undergraduate"
  ) |> 
  relocate(program) |> 
  nest(.by = c(multiple_instructors, last_and_first_name_combined)) |> 
  pmap(process_instructor_names) |> 
  list_rbind() |> 
  mutate(
    across(last_name:first_name, str_to_title),
    # Fix names where title case does not apply
    last_name = case_match(
      last_name,
      "Van De Panne"          ~ "van de Panne",
      "Mochetti De Magalhaes" ~ "Mochetti de Magalhaes",
      "Maclean"               ~ "MacLean",
      "Mcgrenere"             ~ "McGrenere",
      .default = last_name
    ),
    first_name = case_when(
      # Used `count(last_name, first_name)` to find inconsistent first names
      last_name == "Wolfman"  ~ "Steven",
      last_name == "Knorr"    ~ "Edwin Max",
      last_name == "Shepherd" ~ "Frederick Bruce",
      .default = first_name
    )
  )

graduate_courses_clean <- graduate_courses |> 
  clean_names() |> 
  rename(
    course = courses,
    full_name  = instructor_last_name,
    enrolment = enrollment
  ) |> 
  mutate(
    across(c(session, department), str_to_upper),
    section = as.character(section),
    # Used `count(full_name)` to find inconsistent names
    full_name = case_match(
      full_name,
      "Mark Schmidt,CLUNE, JEFFREY" ~ "Mark Schmidt;Jeffrey Clune",
      "Jeff Clune"                  ~ "Jeffrey Clune",
      "Park, Mijung"                ~ "Mi Jung Park",
      "Alex Summers"                ~ "Alexander Summers",
      "Heather O'Brian"             ~ "Heather O'Brien",
      "Heather O’Brien"             ~ "Heather O'Brien",
      "Laks Laksmanan"              ~ "Laks Lakshmanan",
      "Karon Maclean"               ~ "Karon MacLean",
      "Frank Wood"                  ~ "Frank Donald Wood",
      "Bruce Shepherd"              ~ "Frederick Bruce Shepherd",
      "Robert Xiao"                 ~ "Bo Robert Xiao",
      .default = full_name
    ),
    program = "graduate"
  ) |> 
  separate_longer_delim(full_name, delim = regex(";"))

fixed_names <- undergraduate_courses_clean |> 
  select(last_name, first_name) |> 
  distinct() |> 
  # Instructor names on `grad` sheet follow `<first> <last>` format
  mutate(full_name = str_c(first_name, last_name, sep = " "))

fixed_names <- 
  # Instructors only in graduate courses
  graduate_courses_clean |> 
  select(full_name) |> 
  distinct() |> 
  anti_join(fixed_names, by = join_by(full_name)) |> 
  separate_wider_delim(
    full_name,
    names = c("first_name", "last_name"),
    delim = " ",
    cols_remove = FALSE) |>
  bind_rows(fixed_names)
  
graduate_courses_clean <- graduate_courses_clean |>
  inner_join(fixed_names, by = join_by(full_name)) |> 
  select(!c(full_name, instructor_first_name))
```

### Calculate percentage of courses taught by instructors

```{r}
#| label: calculate-percentage

all_courses <- undergraduate_courses_clean |> 
  bind_rows(graduate_courses_clean) |> 
  mutate(
    percentage = 1 / n(),
    .by = year:section
  )

teaching_by_instructor <- all_courses |> 
  summarize(
    total_percentage = sum(percentage),
    .by = c(last_name, first_name)
  ) |> 
  arrange(last_name, first_name)

teaching_by_instructor_and_year <- all_courses |> 
  summarize(
    total_percentage = sum(percentage),
    .by = c(last_name, first_name, year)
  ) |> 
  arrange(last_name, first_name, year)

undergraduate_courses_clean <- all_courses |> 
  filter(program == "undergraduate") |> 
  select(!program)

graduate_courses_clean <- all_courses |> 
  filter(program == "graduate") |> 
  select(!program)
```

### Quality check

The number of undergraduate and graduate courses should be the same as the sum of percentages of courses taught by all instructors.

```{r}
#| label: quality-check

nrow(undergraduate_courses) + nrow(graduate_courses) ==
  sum(all_courses$percentage)
```

## Save data

```{r}
#| label: save-data

write_xlsx(
  list(
    combined = all_courses,
    teaching_by_instructor = teaching_by_instructor,
    teaching_by_instructor_and_year = teaching_by_instructor_and_year,
    undergrad = undergraduate_courses_clean,
    grad = graduate_courses_clean
  ),
  path = here("data", "processed", "course_counts_clean.xlsx")
)
```
